Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs

### 2019-10-04 manuscript/manuscript.Rmd
hagy response done, minus lab methods (still waiting on deets)

Lake datasets that cover longer time periods and broader spatial scales are now becoming available.  Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis. These data allow for analyses that can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing]. Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental]. Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking]. These trophic state models indicate that landscape variables (i.e. ecoregion, elevation, and latitude) are important and that regional trends exist. Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance]. Despite these challenges, it is important to study lakes at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long].

Previous studies using regional data from the northeastern and midwestern United States (US) have investigated spatial and temporal water quality trends and have shown differences based on scale. Macro-scale (i.e. subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale]. This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the individual nutrients [@collins2017lake]. On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence]. Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017]. Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality in Rhode Island lakes and reservoirs.

Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).   URIWW is a citizen science program founded in the late 1980's that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.  Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season.  These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing]. Here specifically, URIWW data contributed to not only the larger regional study by Oliver et al. [-@Oliver_2017], but also allowed us to examine the long-term trends specifically in Rhode Island.

@@

Additionally, prior studies have looked at water quality trends across a larger region of the northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].  The authors found little change in water quality trends across this region [@Oliver_2017] .  We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al. [-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].  Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.

@@

Anomaly analysis works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.  To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period [e.g., @jones1996calculating].  The resultant value is commonly referred to as a z-score.  We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.   Furthermore, since we are interested in water quality trends over time at individual sites, z-scores were calculated over the reference period, 1993-2016, for each site.  We refer to this approach as "site-specific z-scores".

@@

Our analysis indicates that even when nutrient regimes exhibit relative stability (i.e. neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can occur.  Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.  This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.  Also, scale does indeed matter when trying to identify long-term water quality trends. Similar to the results of Oliver et al. [-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g. northeastern and mid-western United States).  However, at the local scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.

@@

As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.  Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent.  However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in chlorophyll *a*  Additionally, these results point to the fact that chlorophyll *a* and aglal biomass is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.

>>>>

Lake datasets that cover longer time periods and broader spatial scales are now becoming available.  Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis. These data allow for analyses that can be useful for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing]. Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental]. Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking]. These trophic state models indicate that landscape variables (e.g. ecoregion, elevation, and latitude) are important and that regional trends exist. Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance]. Despite these challenges, it is important to study lakes at multiple spatial scales because emergent trends on regional or continental scales may or may not be present in individual lakes [@cheruvelil2013multi; @lottig2014long].

Previous studies using regional data from the northeastern and midwestern United States (US) have investigated spatial and temporal water quality trends and have shown differences based on scale. Macro-scale (i.e. subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale]. This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the individual nutrients [@collins2017lake]. On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence]. Thus, it was surprising when little change was reported over a 25 year period for these regions [@Oliver_2017]. Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality in Rhode Island lakes and reservoirs.

Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).   URIWW is a citizen science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.  Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season (e.g. May through October).  These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing]. Here specifically, URIWW data contributed to not only the larger regional study by Oliver et al. [-@Oliver_2017], but also allowed us to examine the long-term trends specifically in Rhode Island.

@@

Additionally, prior studies have looked at water quality trends across a larger region of the northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017]. We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al. [-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].  Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.

@@

Analysis that focuses on anomalies works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.  To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period [e.g., @jones1996calculating].  The resultant value is commonly referred to as a z-score.  We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.   Furthermore, since we are interested in water quality trends over time at individual sites, z-scores were calculated over the reference period, 1993-2016, for each site.  We refer to this approach as "site-specific z-scores".

@@

Our analysis indicates that even when nutrient regimes exhibit relative stability (i.e. neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can can occur.  Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.  This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.  Also, scale does indeed matter when trying to identify long-term water quality trends. Similar to the results of Oliver et al. [-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g. northeastern and mid-western United States).  However, at the local scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.

@@

As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.  Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995, @litke1999review].  However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in chlorophyll *a*  Additionally, these results point to the fact that chlorophyll *a* and aglal biomass is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.
